Evaluating the Importance of Initial Conditions for Antarctic Sea Ice Seasonal Predictability with a Fully Coupled Model

### 2025-10-23 analysis/paper/access-ice.qmd
Adds zenodo data management

execute: 
  echo: false
  warning: false
  message: false
  cache: true

@@

@@

```{r hindcast_extent}
hindcast_extent <- fread(here::here("data/derived/hindcast_extent.csv")) |>
  _[, model := relevel(factor(model), "S2")] |>
  _[, lag := as.numeric(as.Date(time) - as.Date(forecast_time))] |>
  _[lag > 0] |>
  _[!(month(forecast_time) == 1 & member == 7 & model == "S1")] # This member is wrong
```

@@

```{r hindcast_clim}
zeropad <- function(x, n = 2) formatC(x, width = 2, flag = "0")
monnb <- function(d) {
  lt <- as.POSIXlt(as.Date(d, origin = "1900-01-01"))
  lt$year * 12 + lt$mon
}
mondf <- function(d1, d2) {
  monnb(d2) - monnb(d1)
}

months <- 1:12

hindcast_clim <- lapply(setNames(months, months), \(month) {
  lapply(c("S1", "S2"), \(model) {
    here::here(glue::glue(
      "data/derived/climatology/{model}/{zeropad(month)}/em.nc"
    )) |>
      cdo_selmonth(month) |>
      cdo_execute() |>
      ReadNetCDF("aice") |>
      _[, model := model]
  }) |>
    rbindlist()
}) |>
  rbindlist(idcol = "forecast_month") |>
  _[, forecast_month := as.numeric(forecast_month)]

hindcast_clim <- hindcast_clim |>
  copy() |>
  _[,
    .(aice = mean(aice)),
    by = .(month(time), ygrid, xgrid, forecast_month, model)
  ] |>
  _[forecast_month == month]

@@

aice_forecast_example <- c(
  lapply(c("S1", "S2"), \(model) {
    month_clim <- formatC(month(forecast_date), width = 2, flag = "0")

    hindcast(forecast_date, model, 1) |>
      cdo_ydaysub(here::here(glue::glue(
        "data/derived/climatology/{model}/{month_clim}/em.nc"
      ))) |>
      cdo_execute() |>
      ReadNetCDF("aice", subset = list(time = sample_date)) |>
      _[, model := model]
  }),
  list(
    CDR() |>
      anomalies() |>
      ReadNetCDF("aice", subset = list(time = sample_date)) |>
      _[, model := "cdr"]
  )
) |>
  rbindlist()

thickness_example <- c(
  S2 = "/g/data/ux62/access-s2/hindcast/raw_model/ice/hi/daily/e01/di_hi_{format(forecast_date, '%Y%m%d')}_e01.nc",
  S1 = "/g/data/ub7/access-s1/hc/raw_model/unchecked/ice/ice/daily/e01/di_ice_{format(forecast_date, '%Y%m%d')}_e01.nc"
) |>
  purrr::imap(\(f, model) {
    f |>
      glue::glue() |>
      cdo_selname("hi") |>
      remap_cdr() |>
      cdo_execute(cache = TRUE) |>
      ReadNetCDF("hi", subset = list(time = sample_date))
  }) |>
  rbindlist(idcol = "model")

@@

mean_thickness <- function(concentration, thickness, cell_area = NULL) {
  if (is.null(cell_area)) {
    cell_area <- concentration |>
      rcdo::cdo_gridarea()
  }

  volume <- thickness |>
    rcdo::cdo_mul(concentration) |>
    rcdo::cdo_mul(cell_area) |>
    rcdo::cdo_fldsum()

  area <- concentration |>
    rcdo::cdo_mul(cell_area) |>
    rcdo::cdo_fldsum()

  rcdo::cdo_div(volume, area) |>
    rcdo::cdo_execute(options = "-L")
}
mean_thickness_access <- \(forecast_time, model) {
  concentration <- hindcast(
    forecast_time,
    model = model,
    members = "em",
    variable = "aice"
  )
  thickness <- hindcast(
    forecast_time,
    model = model,
    members = "em",
    variable = "hi"
  )

  if (length(thickness) == 0) {
    return(NA_character_)
  }
  if (length(concentration) == 0) {
    return(NA_character_)
  }

  rcdo::cdo_cache_set(here::here("data/temp/cache/"))
  cell_area <- thickness |>
    rcdo::cdo_gridarea()

  mean_thickness(concentration, thickness, cell_area)
}
```

```{r hindcast_thickness}
hindcast_thickness <- CJ(model = c("S2", "S1")) |>
  _[, .(forecast_time = get_forecast_times(model)), by = model] |>
  _[mday(forecast_time) == 1] |>
  _[,
    thickness := unlist(mirai_map(
      data.frame(forecast_time, model),
      mean_thickness_access,
      hindcast = hindcast,
      mean_thickness = mean_thickness
    )[.progress])
  ] |>
  na.omit()

@@

```{r hindcast_mean_thickness}
hindcast_mean_thickness <- hindcast_thickness |>
  _[,
    thickness |> cdo_monmean() |> cdo_execute(),
    by = .(model, forecast_time)
  ] |>
  _[,
    metR::ReadNetCDF(V1, c(thickness = "hi")),
    by = .(model, forecast_time)
  ] |>
  _[, time := as.Date(update(time, day = 1))] |>
  _[, lag := months_difference(time, forecast_time)]
```

@@

```{r errors}
errors <- rbind(
  readRDS(here::here("data/derived/rmse.Rds")),
  readRDS(here::here("data/derived/iiee.Rds"))
) |>
  _[value < 0.01, value := NA] |>
  _[value > 3e13, value := NA] |>
  _[, time := as.Date(time)] |>
  _[, lag := as.numeric(time - time_forecast)] |>
  _[version == "persistence", member := "0em"] |>
  _[member == "0em"] |>
  _[lag > 0] |>
  _[!(month(time_forecast) == 1 & member == "07" & version == "S1")] # This member is wrong
```

@@

```{r topo_lonlat}
topo_lonlat <- here::here("data/raw/ETOPO.nc") |>
  ReadNetCDF() |>
  na.omit()
```

@@

```{r rmse_lon_mean}
rmse_lon_mean <- here::here("data/derived/rmse_lon.Rds") |>
  readRDS() |>
  _[version == "persistence", member := "0em"] |>
  _[member == "0em"] |>
  _[, lon := lon + 12] |>
  _[, time := update(time, hour = 0)] |>
  _[, time := as.Date(time)]
```

@@

```{r errors_ensemble} 
errors_ensemble <- readRDS(here::here("data/derived/rmse.Rds")) |>
  _[version != "persistence"] |>
  _[member != "0em"] |>
  _[value < 0.01, value := NA] |>
  _[value > 3e13, value := NA] |>
  _[!is.na(value)] |>
  _[, time := as.Date(time)] |>
  _[, lag := as.numeric(time - time_forecast)] |>
  _[lag > 0] |>
  _[!(month(time_forecast) == 1 & member == "07" & version == "S1")]

>>>>

execute: 
  echo: false
  warning: false
  message: false
  cache: refresh

@@

```{r zenodo, cache = FALSE}
source(here::here("R/zenodo.R"))
zenodo_init()
zenodo_download_data()
```

@@

```{r hindcast_extent}
hindcast_extent <- here::here("data/derived/hindcast_extent.csv") |>
  zenodo() |>
  fread() |>
  _[, model := relevel(factor(model), "S2")] |>
  _[, lag := as.numeric(as.Date(time) - as.Date(forecast_time))] |>
  _[lag > 0] |>
  _[!(month(forecast_time) == 1 & member == 7 & model == "S1")] # This member is wrong
```

@@

```{r hindcast_clim}

hindcast_clim <- hindcast_0lag_clim() |>
  _[, ReadNetCDF(file, "aice"), by = .(model)] |>
  _[, month := month(time)] |>
  _[, time := NULL]

monnb <- function(d) {
  lt <- as.POSIXlt(as.Date(d, origin = "1900-01-01"))
  lt$year * 12 + lt$mon
}
mondf <- function(d1, d2) {
  monnb(d2) - monnb(d1)
}

@@

aice_forecast_example <- c(
  lapply(c("S1", "S2"), \(model) {
    month_clim <- formatC(month(forecast_date), width = 2, flag = "0")
    clim <- here::here(glue::glue(
      "data/derived/climatology/{model}/{month_clim}/em.nc"
    )) |>
      zenodo()
    hindcast(forecast_date, model, 1) |>
      cdo_ydaysub(clim) |>
      cdo_execute() |>
      ReadNetCDF("aice", subset = list(time = sample_date)) |>
      _[, model := model]
  }),
  list(
    CDR() |>
      anomalies() |>
      ReadNetCDF("aice", subset = list(time = sample_date)) |>
      _[, model := "cdr"]
  )
) |>
  rbindlist()

thickness_example <- c(
  S2 = "/g/data/ux62/access-s2/hindcast/raw_model/ice/hi/daily/e01/di_hi_{format(forecast_date, '%Y%m%d')}_e01.nc",
  S1 = "/g/data/ub7/access-s1/hc/raw_model/unchecked/ice/ice/daily/e01/di_ice_{format(forecast_date, '%Y%m%d')}_e01.nc"
) |>
  zenodo() |>
  purrr::imap(\(f, model) {
    f |>
      glue::glue() |>
      cdo_selname("hi") |>
      remap_cdr() |>
      cdo_execute(cache = TRUE) |>
      ReadNetCDF("hi", subset = list(time = sample_date))
  }) |>
  rbindlist(idcol = "model")

@@

# mean_thickness <- function(concentration, thickness, cell_area = NULL) {
#   rcdo::cdo_cache_set(here::here("data/temp/cache/"))
#   if (is.null(cell_area)) {
#     cell_area <- thickness |>
#       rcdo::cdo_gridarea()
#   }

#   volume <- thickness |>
#     rcdo::cdo_mul(concentration) |>
#     rcdo::cdo_mul(cell_area) |>
#     rcdo::cdo_fldsum()

#   area <- concentration |>
#     rcdo::cdo_mul(cell_area) |>
#     rcdo::cdo_fldsum()

#   rcdo::cdo_div(volume, area) |>
#     rcdo::cdo_execute(options = "-L")
# }

# mean_thickness_access <- \(forecast_time, model) {
#   concentration <- hindcast(
#     forecast_time,
#     model = model,
#     members = "em",
#     variable = "aice"
#   )
#   thickness <- hindcast(
#     forecast_time,
#     model = model,
#     members = "em",
#     variable = "hi"
#   )

#   if (length(thickness) == 0) {
#     return(NA_character_)
#   }
#   if (length(concentration) == 0) {
#     return(NA_character_)
#   }

#   cell_area <- thickness |>
#     rcdo::cdo_gridarea()

#   mean_thickness(concentration, thickness, cell_area)
# }
```

```{r hindcast_thickness}
# hindcast_thickness <- CJ(model = c("S2", "S1")) |>
#   _[, .(forecast_time = get_forecast_times(model)), by = model] |>
#   _[mday(forecast_time) == 1] |>
#   _[,
#     concentration := hindcast(
#       forecast_time,
#       model,
#       members = "em",
#       variable = "aice"
#     ) |>
#       zenodo(),
#     by = .(model)
#   ] |>
#   _[mday(forecast_time) == 1] |>
#   _[,
#     thickness := hindcast(
#       forecast_time,
#       model,
#       members = "em",
#       variable = "hi"
#     ) |>
#       zenodo(),
#     by = .(model)
#   ] |>
#   na.omit() |>
#   _[,
#     thickness := unlist(mirai_map(
#       data.frame(concentration, thickness),
#       mean_thickness
#     )[.progress])
#   ] |>
#   na.omit()

@@

```{r hindcast_mean_thickness}
hindcast_mean_thickness <- hindcast_mean_thickness() |>
  _[,
    ReadNetCDF(mean_thickness, c(thickness = "hi")),
    by = .(model, forecast_time)
  ] |>
  _[, time := as.Date(update(time, day = 1))] |>
  _[, lag := months_difference(time, forecast_time)]
```

@@

```{r errors}
errors <- here::here("data/derived/rmse.Rds") |>
  zenodo() |>
  readRDS() |>
  _[value < 0.01, value := NA] |>
  _[value > 3e13, value := NA] |>
  _[, time := as.Date(time)] |>
  _[, lag := as.numeric(time - time_forecast)] |>
  _[version == "persistence", member := "0em"] |>
  _[member == "0em"] |>
  _[lag > 0] |>
  _[!(month(time_forecast) == 1 & member == "07" & version == "S1")] # This member is wrong
```

@@

```{r topo_lonlat}
topo_lonlat <- here::here("data/raw/ETOPO.nc") |>
  zenodo() |>
  ReadNetCDF() |>
  na.omit()
```

@@

```{r rmse_lon_mean}
rmse_lon_mean <- here::here("data/derived/rmse_lon.Rds") |>
  zenodo() |>
  readRDS() |>
  _[version == "persistence", member := "0em"] |>
  _[member == "0em"] |>
  _[, lon := lon + 12] |>
  _[, time := update(time, hour = 0)] |>
  _[, time := as.Date(time)]
```

@@

```{r errors_ensemble} 
errors_ensemble <- here::here("data/derived/rmse.Rds") |>
  zenodo() |>
  readRDS() |>
  _[version != "persistence"] |>
  _[member != "0em"] |>
  _[value < 0.01, value := NA] |>
  _[value > 3e13, value := NA] |>
  _[!is.na(value)] |>
  _[, time := as.Date(time)] |>
  _[, lag := as.numeric(time - time_forecast)] |>
  _[lag > 0] |>
  _[!(month(time_forecast) == 1 & member == "07" & version == "S1")]
